---
title: "microkit-20260324"
author: "diana baetscher"
date: "2026-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I'm analyzing the harbor porpoise data that Jamie generated on a MiSeq single-end 300 bp microkit.

There was 60% PhiX aligned, despite using a more stringent gel extraction (>600 bp) and quantifying the libraries using both the Kapa qPCR kit for Illumina and the Qubit.

Based on the last microkit run, I'll start with the forward primers since those contained usable sequence data previously.

The first few steps are outside of R, using cutadapt to trim primers.

```{sh, eval = F}
# preliminaries to get cutadapt going
(base) AKCJM140-DB23:cutadapt_mismatches diana.baetscher$ conda activate cutadaptenv

DATA=/Users/diana.baetscher/git-repos/HAPO/data/20260324_HAPOeDNA/rawdata

# below, the first set of () creates an array, containing n elements of the desired trimmed and unique names
NAMELIST=$(ls ${DATA} | sed 's/e*_L001.*//' | uniq)
echo "${NAMELIST}"

```

```{sh, eval=F}
# cutadapt remove primers
# single-end, unsure of which side we're sequencing... begin with the forward primer, which has the `-g` for indicating it is the 5' end, not the 3' end:
for i in ${NAMELIST}; do
	cutadapt --discard-untrimmed -g TACTCCTTGAAAAAGCCCATTGTA -o trimmed/${i}_R1.fastq.gz "$DATA/${i}_L001_R1_001.fastq.gz" >> cutadapt_fwd_out.txt
	done


# then unzip the reads in the /trimmed directory
pigz -d *gz
```

There are some empty files - go ahead and move those to a different directory now.

```{sh move-empty-files, eval = F}
find /Users/diana.baetscher/git-repos/HAPO/data/20260324_HAPOeDNA/trimmed -type f -empty -exec mv {} /Users/diana.baetscher/git-repos/HAPO/data/20260324_HAPOeDNA/trimmed_empty/ \;

```

## Dada2 and ASV analysis

Launching into the R-portion of the analysis in dada2

```{r load-libraries}
library(dada2)
library(tidyverse)
```

```{r}
# file location
path <- "../data/20260324_HAPOeDNA/trimmed"

list.files(path)
```

```{r}
fnFs <- sort(list.files(path, pattern = "_R1.fastq", full.names = TRUE))
#fnRs <- sort(list.files(path, pattern = "_R2.fastq", full.names = TRUE))
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

```{r}
plotQualityProfile(fnFs[1:6])
#plotQualityProfile(fnRs[1:2])
```
Those look much better than the previous run... nice!

```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
names(filtFs) <- sample.names
```


I think I just need to process the Fwd reads
```{r}
# trim reads using the quality scores
out <- filterAndTrim(fnFs, filtFs, truncLen = 200,
              maxN=0, truncQ=10, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE) 

head(out)

# no empty files at this stage.

# tmp <- out %>%
#   as.data.frame() %>%
#   filter(reads.out <1) %>%
#   rownames_to_column(var = "file") %>%
#   select(file) %>%
#   mutate(loc = paste0(path,"/",file)) %>%
#   select(loc)
# 
# empty_files <- tmp$loc

# create a directory for files with no reads
#dir.create("../data/20260122_HAPOeDNA/trimmed/no_reads", showWarnings = FALSE)

# move the files
# file.rename(empty_files,
#             file.path("../data/20260122_HAPOeDNA/trimmed/no_reads", basename(empty_files)))
```


### Error rates
```{r}
my_list <- as.list(filtFs)
new_list <- my_list[1:8] # smaller amount of data for learning the error rates.

# fwd error rates
errF <- learnErrors(new_list, multithread=TRUE)
# plot the erors
p1 <- plotErrors(errF, nominalQ=TRUE)

p1

```

### Sample inference

```{r}
# forwards
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)

```


Make a sequence table
```{r}
seqtab <- makeSequenceTable(dadaFs)
dim(seqtab)

as_data_frame(getSequences(seqtab))
```


```{r}
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
```

All sequences were trimmed to 200 bp per our initial trimming step.

Remove chimeras
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
```

We may decide at a later point that the step to remove chimeras actually removes relevant haplotypes.


## Export files for taxonomy and samples/ASVs

```{r regseqs-asv-output}
 #make fasta file with ASVs
    asv_seqs=colnames(seqtab.nochim)
    for(i in 1:length(asv_seqs))
    {
        write.table(paste(">ASV",i, sep=""),file="csv_outputs/HAPO_micro_20260324_ASV_fwd200bp.csv", append=TRUE, col.names = F, row.names = F, quote=F)
        write.table(paste(asv_seqs[i], sep=""),file="csv_outputs/HAPO_micro_20260324_ASV_fwd200bp.csv", append=TRUE, col.names = F, row.names = F, quote=F)
    }
```

That's the input for the FASTA blastn search.

Goal: change ASV headers to numbered ASVs that correspond to those output in the FASTA file.
```{r output-asv-table}
# Make map between brief names and full sequences
briefToSeq <- colnames(seqtab.nochim)
names(briefToSeq) <- paste0("ASV", seq(ncol(seqtab.nochim))) # Seq1, Seq2, ...
# Make new sequence table with brief names
st.brief <- seqtab.nochim
colnames(st.brief) <- names(briefToSeq)

# export the pool seq table with brief names:
write.csv(st.brief, file="csv_outputs/microkit_20260324_fwd_sampleTable.csv")
```

So a few things to consider:
1) it might be worthwhile to create a custom reference database with harbor porpoise haplotypes for the "blastn" search



I checked the first few ASVs on NCBI and they're all Phocoena phocoena, which is good. So let's treat the ASVs like data for a second.

## Sample info

```{r asvs-in-samples}
tmp <- as.data.frame(st.brief)
sample_df <- rownames_to_column(tmp, var = "sample")

long_df <- sample_df %>%
  pivot_longer(2:length(sample_df)) %>%
  rename(reads = value, ASV = name) 

long_df %>%
  write_csv("csv_outputs/microkit_samples_20260324.csv")

long_df %>%
  filter(reads > 1) %>% # pretty minimal read filter
  ggplot(aes(x = sample, y = reads, fill = ASV)) +
  geom_bar(stat = "identity") +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, size = 6)
  ) +
  scale_y_continuous(expand = c(0,0))

ggsave("pdf_outputs/microkit_20260324_quicklook.png", width = 10, height = 4)
```

